Computing likelihoods under the SMC for a general class of ARGs.

### 2023-11-28 paper.tex
update intro

[TL:DR] Structure the introduction around this great point from Wong et al.:
Note that it is important to distinguish here between structures and models: 
whether an inference method is based on 
heuristics or a rigorous mathematical model is orthogonal to the level 
of detail provided in its estimate. One could heuristically estimate a 
fully precise ARG, or statistically sample a partial, approximate ARG under 
a model such as the coalescent.
What we want to achieve with this paper is to compute a likelihood for 
an approximate (define) ARG under a approximate model of the coalescent. 
Approximate in two senses: SMC + leave out precise recombination time info.
Goal of the introduction: bring across the need to do this.

The underlying genealogical history that relates a sample of individuals back in time
contains all possible information this sample can provide on the history of demography, 
selection and recombination for the population of interest. 
Reconstructing these genealogies with certainty is impossible however. 
As such, in the past the data structure capturing these inheritance paths 
or Ancestral Recombination Graph (ARG), has usually been considered a nuisance 
variable to be integrated out during statistical inference 
\citep{kuhner_2000, griffiths_ancestral_1996}.

%Although the coalescent with recombination best (?) probabilistic description of 
%the events generating an ARG for a given sample, ...

All these: fully precise ARG under the coalescent with recombination.
Computationally prohibitive to explore ARG space:
there are infinitely many ARGs compatible with any genomic dataset, 
and many of these ARGs will have a comparable likelihood \citep{mcvean_approximating_2004}.
By dropping the fully precise requirement that (scalable) ARGs became computationally 
feasible. First ARGwaver.
% ARGweaver, model-based yet approximate (discretised)
% [TO ADD] also argweaver is the first to compute likelihoods for an ARG under 
% the SMC, musn't forget to include this.
Model-based (an approximation of the coalescent, SMC) + approximate in the sense of discretized.
and later really scalable methods:\\

[TL;DR] next step: heuristic methods
Recently, major advancements relying on heuristic rather than model-based approaches
have made ARG inference scalable.
\citep{rasmussen_2014, kelleher_2019, speidel_2019, schaefer_2021, wohns_2022, 
zhang_biobank-scale_2023, zhan_towards_2023}. These methods
have raised the realistic prospect of ARG-based analysis becoming a standard 
part of the population and statistical genetics toolkit.
% however, scalability has impacted our ability to connect these to any model
This has also meant however that
we are now using the term ARG for a varying range of approximate structures
as generated by these different inference methods \citep{wong_a-general_2023}. 
Besides not having a unified output format, the degree of completeness 
in which genetic inheritance from ancestors to 
descendants is documented by each of these methods varies extensively.
This has complicated the ability to connect the output of this diverse range of 
ARG inference methods to a model-based deescription that would allow us to perform  
probabilistic exploration of the ARG space around the point estimate, 
or likelihood-based inference (that does not just consider individual marginal trees).\\

[TL;DR] Specify  different structural approximations:
1/ Recombination is implicit. Recombinations are not modelled or inferred explicitly, 
they can be observed however as tree changes. However a tree change might be the result of one or 
more recombination events.
2/ each of these tools will deal with or resolve uncertainty 
in a different way.\\

[TL;DR] Specify different model-based approximations:
Coalescent with Recombination as the dominant model to describe ancestry.
Through time different approximations. Wiuf and Hein demonstrating a left-to-right 
algorithm. No longer Markovian. This idea used by McVean and Cardin: SMC. 
And followups (SMC', SMC(k)) \citep{mcvean_approximating_2004}.
SMC and relatives are useful: lot of focus however on integrating over 
the distribution of trees to get $N_e(t)$ [ADD REF: PSMC, MSMC]\\

[TL;DR] Wrap up by making clear that we will combine the two types of approximations:
Here, the idea is to use the SMC in a different way that resonates more 
with the true direction of flow of genetic information.
Results in a concrete way to compute likelihoods for *any* inferred ARG 
by integrating out the time to recombination.
and formulate an algorithm 
that is \textit{general} in the sense that it can deal with the various levels of 
completeness that characterise the field's currents attempts [TO DO: make this sound more positive] 
at reconstucting genome-wide genealogies.

>>>>

[TL:DR] Structure the introduction around this great point from Wong et al.:
Note that it is important to distinguish here between structures and models: 
whether an inference method is based on 
heuristics or a rigorous mathematical model is orthogonal to the level 
of detail provided in its estimate. One could heuristically estimate a 
fully precise ARG, or statistically sample a partial, approximate ARG under 
a model such as the coalescent.

1/ ARGs are awesome, or general intro about approximations

model is a mathematical simplification of reality that aims to capture essential aspects 
about what we observe.
For example, the coalescent, ...
Formal way of recording what we observe, results in data structure 
with a certain level of precision.
For example, ARG, ...
The structure capturing the details of these intricately 
interwoven paths of inheritance is referred to as an ancestral recombination graph (ARG)

For a very long time people have assumed that what is meant by an ARG is the 
fully specified genealogical history  
where all events are uniquely positioned both in time
and along the genome. [However see Wong et al. 2023].

Reconstructing these genealogies with certainty about the assumed level of detail 
is impossible.
As such, in the past the data structure capturing these inheritance paths 
or Ancestral Recombination Graph (ARG), has usually been considered a nuisance 
variable to be integrated out during statistical inference 
\citep{kuhner_2000, griffiths_ancestral_1996}.\\

% 2/ approximations have driven the rate of progress in inference
Classical approach Hudson, Markovian backwards in time.
McVean and Cardin, SMC, when described left-to-right also Markovian 
in that sense. By sacrificing precision on the our ability to capture long 
range linkage information gain tractability.

SMC functioned as the engine of many powerful inference methods 
if the continuous state space for the Markov chain 
(consisting of all possible genealogies) is approximated by a moderately 
sized finite set—typically by enumerating tree topologies and/or discretizing 
time—then inference can be performed efficiently using well-known algorithms 
for hidden Markov models (HMMs).
ARGweaver uses these same insights to develop the first real milestone in ARG inference.\\

% 3/ Recent progress in ARGs through approximate structures

% bring nuance in the distinction between detail and approximation

% heuristic methods guided by coalescent principles

% what is an approximate structure: implicit recombination, polytomies or not, ...

These methods
have raised the realistic prospect of ARG-based analysis becoming a standard 
part of the population and statistical genetics toolkit.

% 4/ Big problem: how do you connect such approximate structures each 
% with different levels of detail to models?
These recent developments have also meant however that
we are now using the term ARG for a varying range of approximate structures 
as generated by these different inference methods \citep{wong_a-general_2023}. 
Besides not having a unified output format, the degree of completeness 
in which genetic inheritance from ancestors to 
descendants is documented by each of these methods varies extensively.
This has complicated the ability to connect the output of this diverse range of 
ARG inference methods to a model-based deescription that would allow us to perform  
probabilistic exploration of the ARG space around the point estimate, 
or likelihood-based inference (that does not just consider individual marginal trees).

% 5/ what we will do

Here, the idea is to use the backwards-in-time formulation of the SMC to 
compute likelihoods for for \textit{any} inferred ARG 
by integrating out the time to recombination.
The application of the SMC in a way that resonates more 
with the true direction of flow of genetic information 
results in a concrete way to formulate an algorithm 
that is \textit{general} in the sense that it can deal with the various levels of 
completeness that characterise the current state of the field of ARG reconstruction.
